---
title: Benchmarking sequence, embedding and structure-based methods for eukaryotic virus protein annotation
metadata-files:
  - _authors.yaml
bibliography: references.bib
format:
  nature-pdf:
    notebook-view: false
    notebook-links: false
    journal: 
      cite-style: sn-nature
    cite-method: citeproc
    colorlinks: true
    linkcolor: black
    fig-pos: "H"
    keep-tex: true
    include-in-header:
        - _extensions/fonts.tex
        - text: |
            \usepackage{lineno}
            \linenumbers
            \linespread{1.5}
            \usepackage{pdflscape}
            \usepackage{placeins}
    mermaid-format: png
  docx:
    reference-doc: custom-reference-doc.docx
    filters:
        - authors-block
    notebook-view: false
    notebook-links: false
    mermaid-format: png
csl: _extensions/christopherkenny/nature/csl/nature.csl
notes-after-punctuation: false
execute:
  freeze: auto  # re-render only when source changes
crossref:
  custom:
    - kind: float
      reference-prefix: Supplementary Figure
      key: suppfig
      latex-env: suppfig
      latex-list-of-description: Supplementary Figure
abstract: |
  The advent of protein language models and structural prediction promises new avenues for protein annotation through structure homology. Earlier work (eg. Phold) has shown protein structural homology annotation works well for prokaryotic viruses. To leverage these new techniques for protein annotation of eukaryotic viruses, we benchmarked four recent embedding- and structure-based methods in comparison with the gold standard of sequence-based annotation. We selected 10% of the entries in ICTV's VMR v40.2 that did not have bacteria or archaea as designated host (n=1,054), and subsequently retrieved all associated proteins from NCBI (n=11,360). Boltz-2 structure prediction of these proteins produced generally high-quality models (median pLDDT 71.3). Using the structures and protein sequences of the entries in the Big Fantastic Virus Database, we queried the VMR proteins with BLASTp, Diamond, and MMseqs2 (sequence); TEA and ProstT5-3Di (embedding); and Foldseek and Reseek (structure). Sequence homology-based methods showed a high overlap in identical hits, whereas the other methods had low intra- and inter-method agreement in hits between themselves and the sequence-based methods. Additionally, all methods improved the number of informative categories (ie. not “hypothetical protein”) for the VMR proteins, with embedding and structure methods producing the largest increases. However, when evaluated against proteins with informative GenBank annotations as ground truth (n=3,983), embedding- and structure-based methods agreed less often with the "ground truth" than sequence-based methods (maximum 63.6% and 72.2% vs 78.4%, respectively). Overall, sequence homology remains the most accurate approach for protein annotation of eukaryotic viruses at present, although for a minority of "hypothetical proteins" in mostly large dsDNA viruses, structure homology could help resolve functional annotation.
---

# Introduction
The current increase in metagenomic sequencing experiments is providing researchers with a steady stream of new viral genomes. A large problem that remains for computational biologists is the annotation of these genomes with functionally relevant labels for the predicted proteins. As annotating all these genomes through experimental evidence is an unfeasible task, researchers transfer annotations of annotated genomes to these novel genomes. The established gold standard for performing this task has been the use of the BLAST suite or comparable sequence alignment tools, such as Diamond and MMSeqs2 [@altschulBasicLocalAlignment1990; @buchfinkFastSensitiveProtein2014; @steineggerMMseqs2EnablesSensitive2017]. However, this approach reaches its limits when newly discovered sequences diverge too much from experimentally validated annotated genomes, a problem that is particularly acute in virology, where mutation rates are high and sequence diversity is vast @simmondsVirusTaxonomyAge2017.

Protein language models (pLMs) have emerged as a powerful paradigm for capturing biologically meaningful representations from amino acid sequences alone, trained on known protein sequences using self-supervised objectives analogous to those used in natural language processing [@rivesBiologicalStructureFunction2021; @elnaggarProtTransUnderstandingLanguage2022]. By learning the statistical regularities of protein sequences across millions of examples, pLMs implicitly encode evolutionary, structural, and functional information in dense embedding vectors. These representations have proven useful for a wide range of downstream tasks, including secondary structure prediction, contact map prediction, solubility prediction, and functional site annotation [@rivesBiologicalStructureFunction2021; @elnaggarProtTransUnderstandingLanguage2022; @linEvolutionaryscalePredictionAtomiclevel2023]. Models such as ESM-2 [@linEvolutionaryscalePredictionAtomiclevel2023] and ProtTrans [@elnaggarProtTransUnderstandingLanguage2022] have become widely used as general-purpose sequence encoders in the field.

A complementary and increasingly important line of research leverages protein three-dimensional structure, which is generally more conserved than primary sequence over long evolutionary timescales, because the topology of a folded protein is directly tied to its biochemical function [@illergardStructureThreeTen2009]. The advent of highly accurate structure prediction tools such as AlphaFold2 [@jumperHighlyAccurateProtein2021] and, more recently, Boltz-2 [@passaroBoltz2AccurateEfficient2025], has dramatically lowered the barrier to obtaining structural models. Structural comparison tools, most notably Foldseek [@vankempenFastAccurateProtein2024], have made it feasible to search large structural databases at scale by encoding local backbone geometry into a compact structural alphabet (3Di), enabling alignment-based searches analogous to BLAST but operating on structure rather than sequence. More recently, Reseek [@edgarProteinStructureAlignment2024] was developed to further improve sensitivity to remote structural homologs.

The convergence of pLMs and structural biology has opened a third avenue: using language models to predict structural information without the computational cost of three-dimensional structure prediction. ProstT5 is a notable example, fine-tuning a protein language model to translate amino acid sequences into 3Di token sequences, effectively producing a structural representation with a three order magnitude speedup compared to structure prediction [@heinzingerBilingualLanguageModel2024]. A related approach, The Embedded Alphabet (TEA), uses ESM-2 embeddings as input to generate an alternative structural alphabet that the authors argue better represents helices and loops compared to 3Di [@pantoliniRewritingProteinAlphabets2026]. These compressed one-dimensional structural representations can then be used as inputs to standard alignment tools, bridging the gap between pure sequence methods and full structure comparison pipelines.

To address the annotation bottleneck for phage genomes specifically, Phold [@bourasProteinStructureinformedBacteriophage2026] cleverly combines structural information with a large database of high-quality phage protein predictions. Phold either accepts user-provided structure predictions or uses ProstT5 to convert amino acid sequences into 3Di sequences, which are then aligned against annotated references using Foldseek. Benchmarking showed that Phold improved the annotation of a select set of bacteriophage proteins from approximately 35% to approximately 50% with both ProstT5 and protein structures [@bourasProteinStructureinformedBacteriophage2026]. A similar approach, Baktfold, was subsequently developed by the same group to increase annotation rates in bacterial genomes [@bourasBaktfoldSensitiveProtein2026].

In addition to bacteriophages, also less studied eukaryotic viruses (e.g. insect viruses, nematode viruses and plant viruses) might benefit from such an approach to bring a more functional understanding to metagenomic discoveries of these viruses.  For example, the *Chrysoviridae*, a family infecting fungi, plants, and likely insects, whose four dsRNA genome segments encode four proteins. Two of these proteins — an RNA-dependent RNA polymerase and a capsid protein — have established functions, while the remaining two are of entirely unknown function [@kotta-loizouICTVVirusTaxonomy2020].

Therefore, we benchmarked seven approaches for functional annotation: BLASTp, Diamond and MMseqs2 as gold standard sequence alignment methods; ProstT5 3Di conversion with Foldseek alignment, and ESM-2 to The Embedded Alphabet (TEA) conversion with MMseqs2 alignment as pLM-derived structural alphabet approaches; and finally, full three-dimensional structure prediction of viral proteins with Boltz-2 [@passaroBoltz2AccurateEfficient2025] followed by structural alignment using either Foldseek [@vankempenFastAccurateProtein2024] or Reseek [@edgarProteinStructureAlignment2024]. To evaluate these methods on a representative benchmark, we downloaded 11,360 proteins from 1,000 randomly selected, ICTV-ratified eukaryotic virus species, whose formal taxonomic recognition implies a better degree of annotation than randomly selected viral sequences from GenBank, while also being less comprehensively characterized than the curated entries in RefSeq. For all methods, we used the Big Fantastic Virus Database (BFVD), which contains 351,242 high-quality predicted viral protein structures, as a unified reference database [@kimBFVDLargeRepository2025], ensuring that each annotation method was evaluated against an identical set of viral proteins and that observed differences in performance could be attributed to the annotation strategy rather than to differences in the underlying reference.
 
# Results
## Boltz-2 generates high-quality viral protein structures without parameter fine-tuning
Our test set of protein sequences consisted of 1,054 different viral entries taken at random from ICTV's VMR data (n=1,000 distinct species). These test dataset viruses covered all eukaryotic viral Kingdoms, with the majority of proteins originating from *Orthornavirae* (RNA viruses) and *Bamfordvirae* (dsDNA viruses) (see [@fig-overview]A), despite the latter only representing 43 viruses (see [@suppfig-kingdom]). 

Structure prediction of all 11,467 proteins from our test set with Boltz-2 resulted in high-quality predicted structures, with a median pLDDT score of 71.3 and 52.4% of all predicted structures having a pLDDT score above 70 (see [@fig-overview]B). This shows that the structure prediction with default settings of Boltz-2 results in reliable structure predictions without the need for extensive fine-tuning of parameters, although extra viral sequences were used for the multiple sequence alignment (MSA) generation (see [methods](#structure-prediction)).

![A) Test protein distribution from the different viral Kingdoms. Each tile roughly corresponds to 110 proteins. B) A cumulative distribution function plot showing the percentage of predicted structures from the test set with a pLDDT score above the pLDDT score on the x-axis. The inset shows a violin- and boxplot with all pLDDT scores from the predicted structures (n=11,467).](../figures/data_info/dataset_overview.pdf){#fig-overview}

## Sequence homology methods demonstrate superior sensitivity and reproducibility over structure-informed approaches in eukaryotic virus annotation
We started by evaluating how many of the proteins in our test produced at least one match against the BFVD for each method (see [@fig-overlap]A). In addition, we also filtered the results of each method on an E-value lower than 1e-03 to see how this affected the number of hits for each test protein. This decreased the number of test proteins with a hit on average with 25% for the structure methods. The methods relying on structure homology clearly had less matches compared to the sequence homology-based methods. This suggests that sequence-based methods are still more sensitive in detecting homology for eukaryotic viruses than structure-based methods, despite the promising results for bacteriophages [@bourasProteinStructureinformedBacteriophage2026]. As the quality of our structure predictions was rather high (see [@fig-overview]B), this lower sensitivity is likely not due to poor structure predictions. In contrast, TEA had, as the only non-protein sequence-based method, a comparable number of test proteins with a match, almost reaching the total number of test proteins (n=11,360). However, analyzing the total number of hits at different E-value thresholds shows that the TEA search resulted in a very high number of matches across all E-value thresholds, signifying a high number of false positives (see [@fig-overlap]B). A similar pattern was observed for the other structure-based methods, which tended to find too many hits at high E-values. Reseek was less affected by this, due to a combination of already calculating E-values more accurately [@edgarProteinStructureAlignment2024; @edgarProteinStructureAlignment2025] and our use of a modified Reseek version (see [below](#reseek)). On the other hand, the three sequence-based methods had a similar number of test proteins with a hit (BLASTp=11,359; MMseqs2=11,255 and Diamond=11,224), but don't show this inflation of total hits at higher E-values (BLASTp had a slight increase towards larger E-values).

When comparing the BFVD matches yielded by all methods against each other (see [@fig-overlap]C-D and [@suppfig-pairwise-method]), we observed that the sequence homology-based methods had a high overlap in identical hits. This was true when comparing the best 25 hits for each test protein (only the comparison of diamond and mmseqs2 against blastp had a low overlap because of the higher amount of hits for blastp), the best 25 hits afterwards filtered on E-value, and only the top hits (see [@fig-overlap]C-D and @suppfig-pairwise-method). In contrast, the embedding and structure homology-based methods had low intra-method agreement in resulting hits (~25% between the embedding methods and ~40-75% for the structure methods on the best 25 hits). The percentage of overlapping hits did increase when only taking the top hits into account, but stayed well below the percentage of overlap between the sequence-based methods. It is worth noting that Reseek seemed to perform worse in this metric when looking at the filtered results (see [@suppfig-pairwise-method]B), however this is due to the large reduction in number of matches after filtering. This reduction is due to the more stringent E-value calculations in Reseek compared to Foldseek [@edgarProteinStructureAlignment2024; @edgarProteinStructureAlignment2025]. Notably, while ProstT5 was the worst performer across the three datasets compared to the sequence methods, TEA performed better in the top hits comparison and was on par with the structure-based methods. Overall, the embedding and structure-based methods, that should be better to detect remote homologs, had low agreement on shared hits among each other. 

![A) The number of proteins that got a hit against the BFVD shown per method (colored by category) before and after filtering the reported results on E-value lower than 1e-03 (lighter color). B) The total number of hits across all test proteins in function of the E-value for each method. C-D) Pairwise comparisons of shared hits across methods for the same test proteins, shown for the best 25 hits (C), and the top hits regardless of E-value (D). Text inside the colored tiles indicate the number of shared hits for the method on the x-axis compared to the method on the y-axis, followed by the percentage of shared hits as a proportion of the total hits for the method on the y-axis.](../figures/search_results/hit_overlap_summary.pdf){#fig-overlap}

## Sequence-based methods retain better accuracy in functional category assignment while structural approaches slightly improve novel annotations
Using a weighted approach, we assigned functional categories to the test proteins for each method based on the hits with an E-value smaller than 1e-03 (see [methods](#functional-category-assignment)). To evaluate the accuracy of the benchmarked methods, we considered the test proteins with an originally informative annotation (n=3,983) as the ground truth to compare our results to. All methods improved the number of informative categories (see [@fig-annotation]A), with the sequence-based methods, together with TEA and Boltz-Foldseek, improving the number of informative assignments the most. However, when comparing the functional annotations to the "ground truth", all embedding and structure-based methods underperformed compared to the sequence-based methods (maximum 63.6% and 72.2% vs 78.4% agreement with "ground truth"; see [@fig-annotation]B). 

Although TEA made the largest improvements in assigning previous uninformative proteins to informative categories, it also showed the greatest inconsistencies between assigned informative categories that were already informative in the original test set (see [@fig-annotation]C). In addition, it also has 558 queries that were changed from an informative function to a non-informative one, which is almost double the amount compared to blastp which had the second highest informative-to-uninformative reassignments (see [@fig-annotation]C and @tbl-assigned-categories). The same applies to the number of test proteins that originally had an informative category, but was assigned a different category with TEA (n=1,062). Furthermore, due to the likely high number of false positives in TEA hits (see [@fig-overlap]B), we concluded that the TEA uninformative-to-informative assignments lacked reliability. 

Of note, Boltz structure prediction combined with Foldseek did assign more informative categories than the sequence-based methods to originally non-informative proteins (n=1,578 vs ~1,460; see @tbl-assigned-categories). However, this method retrieved less test proteins with hits against the BFVD. Interestingly, 337 proteins were assigned an informative category with Boltz-foldseek that were uninformative with either of the sequence-based methods. Most of these proteins belonged to the *Bamfordvirae*, a kingdom including nucleocytoplasmic large dsDNA viruses (NCLDVs) characterized by their massive genome sizes. Furthermore, also proteins from the *Heunggongvirae* kingdom were overrepresented compared to their presence in the total dataset (see [@fig-overview]A and [@fig-annotation]D). The eukaryotic viruses in the *Heunggongvirae* consist of members of the *Herpesvirales*, which also have large dsDNA genomes. In addition, the quality of the predicted protein structure was an important factor in the assignment of an informative category by Boltz-foldseek as >90% of these proteins had at least a medium quality structure, while across the structure prediction of all proteins this was 79.3% (see [@fig-overview]B and [@fig-annotation]E). For high-quality structures, this percentage even almost doubled (15% vs 8.7%). Finally, we looked at the categories these proteins belonged to and found that most (~40%) of them belonged to proteins with a structural function (see [@fig-annotation]F; includes both Structural and pvp - phage virion proteins). Also, ~10% of the proteins were assigned an "Anti-viral evasion" function which was mostly driven by Ser/Thr-kinases and interferon antagonist assignments.

![A) The number of proteins that received an informative functional annotation (i.e. not "Hypothetical Protein", "unknown" or "Other Function") after applying each method. Colored by method, light shades indicate non-informative proteins, dark shade indicates informative protein assignments. B) The percentage of queries that have a matching functional category for each method compared tot the informative original functional categories (regarded as the ground truth). C) Comparison of final assigned annotation categories compared to the original categories. Negative changes (ie. different informative categories or informative labels that changed to non-informative categories) are displayed to the left of the zero point, matches or category improvements (from non-informative to informative) are displayed on the right. D-F) Waffle plots showing the proportions of proteins that could be assigned an informative function with Boltz-foldseek and not with the sequence based methods, colored by D) taxonomy on Kingdom level, E) structure quality and F) main activity the assigned functional category takes part in (starts with uppercase; based on @tbl-supp-functional-categories) or the main category of the Empathi hierarchical classification system (starts with lowercase). Tiles are ordered from bottom to top, and from left to right. Each tile corresponds to roughly three proteins.](../figures/search_results/annotation_combined_2.pdf){#fig-annotation}


```{r}
#| label: tbl-assigned-categories
#| tbl-cap: "Functional annotation outcomes across the seven benchmarked methods."
#| warning: false
#| message: false
assigned_categories_path <- "../results/category_assignment_results.tsv"
assigned_categories <- readr::read_tsv(assigned_categories_path) |>
  dplyr::select(
    method,
    queries_with_hit,
    -matched_original_category,
    matched_original_informative,
    matched_original_deprioritized,
    original_informative_changed,
    original_deprioritized_to_informative,
    original_informative_to_deprioritized,
    -pct_match_among_hits,
    -pct_deprioritized_to_informative_among_hits,
    -pct_informative_to_deprioritized_among_hits
  ) |>
  dplyr::rename(
    `Queries-assigned` = queries_with_hit,
    `Informative-match` = matched_original_informative,
    `Uninformative-match` = matched_original_deprioritized,
    `Informative-mismatch` = original_informative_changed,
    `Uninformative-to informative` = original_deprioritized_to_informative,
    `Informative to-uninformative` = original_informative_to_deprioritized,
  ) |>
  dplyr::mutate(
    method = factor(
      method,
      levels = c(
        "blastp",
        "diamond",
        "mmseqs2",
        "ProstT5-foldseek",
        "TEA-mmseqs2",
        "Boltz-foldseek",
        "Boltz-reseek"
      )
    )
  ) |>
  dplyr::arrange(method)

assigned_categories_colnames <- names(assigned_categories)

if (knitr::is_latex_output()) {
  assigned_categories_colnames <- stringr::str_replace_all(
    assigned_categories_colnames,
    "-",
    "\n"
  ) |>
    kableExtra::linebreak(align = "r")
} else {
  assigned_categories_colnames <- stringr::str_replace_all(
    assigned_categories_colnames,
    "-",
    "<br>"
  )
}

knitr::kable(
  assigned_categories,
  col.names = assigned_categories_colnames,
  format = if (knitr::is_latex_output()) "latex" else "pipe",
  booktabs = TRUE,
  escape = F,
  linesep = ""
)
```


## Computational benchmarking
All search methods were benchmarked for runtime and memory usage using Hyperfine [@Peter_hyperfine_2023], and `nvidia-smi` for GPU memory consumption. As expected, the sequence-based methods were much faster and less memory intensive compared to the embedding and structure-based methods (see @fig-computational-benchmark and @suppfig-time-increase). Both Diamond and MMseqs2 finished the search of the >11,000 proteins against the BFVD in under 10 seconds, followed by BLASTp with a runtime of almost two minutes, a ~28x time increase compared to MMseqs2 which was the fastest workflow. The embedding methods took 52 and 93 times longer than MMseqs2 for TEA and the ProstT5 workflow, respectively. Sequence-based methods were generally compatible with consumer-grade hardware. In contrast, the TEA workflow using the full precision ESM2 model required substantially greater GPU VRAM, necessitating specialized hardware, whereas ProstT5 operated within the memory constraints of consumer-grade GPUs. The structural workflows, by contrast, required over 20,000-fold more runtime than the sequence-based tools (see @suppfig-time-increase) and demanded substantial computational resources, exceeding 400 GiB of system memory (CPU) and approximately 80 GiB of GPU memory (see @fig-computational-benchmark). Furthermore, the Reseek structural comparison incurred an additional two-hour runtime, whereas the Foldseek-based structural comparison completed in approximately two minutes.

![Log~10~-transformed runtime in seconds of the different workflows plotted in function of the log~10~-transformed memory usage (in GiB). For multistep workflows the indicated time is a cumulation of all steps (arrows indicate directionality of workflow steps). Workflows are colored by search method, CPU and GPU steps are indicated by circles and triangles respectively. The inset shows the difference in runtime between foldseek and reseek after MSA generation (ColabFold) and structure prediction (Boltz2).](../figures/computational_benchmark/hyperfine_time_vs_memory.pdf){#fig-computational-benchmark}

# Discussion
In this study, we benchmarked several established and novel methods for annotating eukaryotic viral proteins with a function. Our goal was to evaluate these tools using largely default settings, reflecting their usability in high-throughput annotation workflows by experienced researchers without specialized expertise in each method. In this regards, all tools performed well and were easy to run due to their extensive documentation. Most notably, structure prediction with Boltz-2 produced high-quality models without extensive parameter tuning. Aside from expanding the ColabFold MSA with additional viral sequences, all runs were performed using default settings, indicating that generating accurate structures is accessible even to researchers with limited experience in structure prediction.

Although new methods like pLMs and structure homology searches have shown great promise in detecting remote homologs [@edgarProteinStructureAlignment2024; @vankempenFastAccurateProtein2024; @pantoliniRewritingProteinAlphabets2026], including bacteriophage protein searches [@bourasProteinStructureinformedBacteriophage2026], our benchmark suggests that these advantages do not yet translate straightforwardly to eukaryotic virus proteins. Across our dataset, the sequence-based methods were consistently the most reproducible (high overlap of identical hits) and the most accurate when evaluated against the original classifications, whereas embedding- and structure-based methods showed lower agreement with the ground truth and substantially lower concordance with one another.

A possible explanation for the low performance of the pLM approaches used here could be the lack of (eukaryotic virus) training data that was used to train these models due to an underrepresentation of viruses in large protein databases (e.g. UniRef90) and biosafety concerns [@panPredictingEvolutionaryFunctional2026; @varadiAlphaFoldProteinStructure2024; @sawhneyFinetuningProteinLanguage2025]. The use of virus-specific pLMs, like LucaVirus [@panPredictingEvolutionaryFunctional2026], could potentially mitigate this issue. Although the LucaVirus pLM is already available to generate the embeddings, their newly developed tool, LMalign, to align the generated embeddings is not. This did not readily allow us to compare the test protein embeddings with embeddings from the BFVD, our unified reference to ensure a fair comparison. On the other hand, it should be noted that recent work on mutational effect prediction with fine-tuned pLMs on viral sequences found that fine-tuning pLMs did not completely resolve the observed performance issues of pLMs on viral datasets [@vieiraIntrinsicDatasetFeatures2026].

In contrast, the structure-based comparison with Foldseek outperformed the best pLM approach by ~9% when benchmarked against the ground truth, yet remained less accurate than sequence-based methods. Despite this limitation, structural comparison enabled the functional annotation of 337 previously hypothetical proteins that were not classified based on  sequence approaches. However, this gain is modest relative to the broader improvements in functional annotation reported for bacteriophages using structure homology [@bourasProteinStructureinformedBacteriophage2026]. The majority of these newly annotated proteins originate from large dsDNA viruses and were assigned structural functions, which likely reflects the overrepresentation of structurally annotated proteins in the BFVD (see @suppfig-bfvd-classification), thereby facilitating the detection of homologs among hypothetical proteins. Notably, approximately 10% of the hypothetical proteins receiving informative annotations were assigned the "Anti-viral evasion" label. Although this category is well-represented in the BFVD, its frequency among newly annotated proteins exceeds what would be expected relative to their presence in the BFVD. This implies that structure-based approaches may be particularly valuable for uncovering anti-viral evasion functions in large dsDNA viruses, consistent with evidence of conserved evasion strategies shared across phylogenetically unrelated virus families such as *Poxviridae* and *Paramyxoviridae* [@talbot-cooperPoxvirusesParamyxovirusesUse2022].

A few limitations arise from our approach: First, polyproteins — large precursor proteins that are subsequently cleaved by proteases into multiple smaller, functional proteins — were treated as single entities throughout our analysis. While this does not accurately reflect their biological reality, it was an inherent constraint imposed by the BFVD and UniProt, neither of which consistently provides information on individual subproteins within a polyprotein. Secondly, our LLM-based classification is susceptible to ambiguity when viral proteins with a distinct function share the same name. A notable example is VP1 (viral protein 1), which designates a capsid protein in *Parvoviridae* [@cotmoreICTVVirusTaxonomy2019] but corresponds to the RNA-dependent RNA polymerase in *Sedoreoviridae* [@matthijnssensICTVVirusTaxonomy2022]. To mitigate this, we provided taxonomic information to the LLM to guide classification; however, classification errors were not completely avoided. For instance, when the name was spelled out in full as "viral protein 1", the LLM failed to correctly classify the *Parvoviridae* VP1, instead assigning it the "Hypothetical protein" label. Nevertheless, even with these shortcomings, the LLM-based approach outperforms classical pattern-based matching in handling such ambiguities. Finally, the structural homology search was constrained by the composition of the BFVD, which currently contains approximately 350,000 structures, a small fraction of all known viral proteins. Nevertheless, the BFVD represents the largest available repository of eukaryotic virus protein structures, and expanding its coverage would represent a major undertaking, requiring substantial computational resources and time.

Overall, our benchmark supports a conservative conclusion: for eukaryotic virus protein annotation, sequence homology remains the most accurate and reproducible primary strategy. Structure-based methods serve as valuable complementary tools; however, their application should be deliberate and targeted given the substantial computational resources they require. The embedding-based methods and pLMs evaluated in this study do not offer sufficient accuracy for reliable functional annotation of eukaryotic virus proteins. While fine-tuned models may improve performance, their accuracy for functional annotation must be rigorously and continuously validated rather than assumed.

# Material and Methods
## Dataset selection
### Test set
To benchmark different methods for eukaryotic virus protein annotation, we selected 10% of the entries in ICTV's Virus Metadata Resource (VMR) v40.2 that did not have bacteria or archaea as designated host (n=1,054; 1,000 distinct species). We subsequently extracted their GenBank accession numbers (n=1,402) and downloaded all associated proteins (n=11,360). 

Protein sequences that were longer than 2,500 amino acids were split into multiple chunks of maximum 2,500 amino acids so that structure prediction with Boltz-2 @passaroBoltz2AccurateEfficient2025 would not run out of GPU memory (see [below](#computational-benchmarking)). This resulted in a final test set of 11,467 protein sequences from 1,000 eukaryotic virus species.

### Database
The BFVD [@kimBFVDLargeRepository2025] was used as the reference database for both sequence and structure homology-based annotation. The BFVD is a comprehensive database of high-quality predicted protein structures from the viral sequence representatives of the UniRef30 clusters. UniRef30 is a UniProt database that clusters protein sequences at 30% sequence identity and selects a representative sequence for each cluster; these representatives thus have a UniProt accession. Since a substantial amount of BFVD's included proteins are no longer available in UniProtKB as they are not part of reference proteomes, we converted their UniProt accessions to UniParc IDs. 

The BFVD database was further converted into the required formats of the search tools used in this benchmarking effort (see [sequence](#sequence-homology-based-annotation) and [structure](#structure-homology-based-annotation) homology-based annotation).

## Functional classifier
Functional categories for each BFVD entry were derived from protein names associated with the corresponding UniParc IDs (could be from UniProtKB/TrEMBL, RefSeq and/or GenBank) for eukaryotic viruses or by classifying them into PHROG-like [@terzianPHROGFamiliesProkaryotic2021] categories with Empathi [@boulayEmpathiEmbeddingbasedPhage2025] in the case of bacteriophage proteins. The distinction between eukaryotic virus and bacteriophage proteins was made based on the TaxID information from the UniParc protein ID. 

Functional categories for eukaryotic virus proteins were designed based on previously established enzymes by ViralZone (see https://viralzone.expasy.org/10417) and adding structural proteins together with extra enzymatic categories. These extra categories were acquired by prompting Perplexity AI in deep research mode with: 

> "Can you give me an exhaustive list of common protein functions in eukaryotic viruses? Use https://viralzone.expasy.org and https://ictv.global/report/genome as sources. Give me a machine readable file with all functional categories." 

The resulting functional categories were further manually curated to ensure that they were mutually exclusive. The final list of functional categories is provided in @tbl-supp-functional-categories. A custom python script was used to assign the established functional categories to the eukaryotic virus BFVD proteins using Google's Gemma 4 31 billion parameter, instruction-trained LLM model locally (https://huggingface.co/google/gemma-4-31B-it). The LLM had access to the collection of protein names and the viral taxonomy of each entry, and was asked to provide a category classification based on this information. Assigned functional categories were checked against the originally established functional categories and removed if they did not correspond.

The same functional classifier script was also used on the protein names from our test set to assign functional categories to the eukaryotic virus proteins downloaded from the ICTV exemplar species.

An overview of the LLM pipeline is given in @suppfig-llm-diagram.

## Sequence homology-based annotation
For sequence homology annotation, we used three state-of-the-art tools widely used in functional annotation pipelines and sequence similarity searches. As gold standard, BLASTp v2.16.0+ [@altschulBasicLocalAlignment1990] was used to align the test set amino acid sequences against the amino acid sequences of the BFVD entries. In addition, we used `diamond blastp` (Diamond v2.1.16) [@buchfinkSensitiveProteinAlignments2021] in sensitive mode and `mmseqs easy-search` (MMseqs2 v18.8cc5c) @steineggerMMseqs2EnablesSensitive2017 with default settings; except for an E-value cutoff of 10 instead of 1e-03 to allow comparison to the other methods.

## Protein embedding-based annotation
### ProstT5
ProstT5 [@heinzingerBilingualLanguageModel2024] is a pLM, based on ProtT5-XL-U50 [@elnaggarProtTransUnderstandingLanguage2022], finetuned for translating amino acid sequences into structure representations with 3Di sequences. We used ProstT5 within foldseek v10.941cd33 [@vankempenFastAccurateProtein2024] using default settings of `foldseek createdb` to generate 3Di sequences for our test set and search them against the BFVD with `foldseek easy-search`.

### TEA
In contrast to ProstT5, TEA [@pantoliniRewritingProteinAlphabets2026] uses the ESM2 pLM [@linEvolutionaryscalePredictionAtomiclevel2023] to generate embeddings for protein sequences and converts these embeddings into TEA sequences using this novel structural alphabet. Despite TEA and 3Di both consisting of 20 characters, TEA is different from 3Di in terms of the numbers and distribution these TEA characters map to secondary structural elements, promising a better representation of different structural elements across multiple alphabet characters. 

We converted both our test set and the BFVD into TEA sequences using the `tea_convert` pipeline with default settings, and subsequently performed sequence similarity searches using `mmseqs easy-search` @steineggerMMseqs2EnablesSensitive2017 with a custom substitution matrix (matcha substiution matrix provided with TEA) and gap scoring settings (`--comp-bias-corr 0 --mask 0 --gap-open 18 --gap-extend 3`) to search the test set against the BFVD.

## Structure homology-based annotation
### Structure prediction
For the structure prediction of our test set proteins, we relied on Boltz v2.2.1 @passaroBoltz2AccurateEfficient2025 (default settings) with MSAs generated by a local, adapted ColabFold installation [@mirditaColabFoldMakingProtein2022] (see <https://github.com/gbouras13/colabfoldv>). To increase the depth and diversity of the MSAs, we used the default ColabFold sequence databases, supplemented with extra *Orthornavirae* sequences [@mutzProteinStructuromeOrthornavirae2024] and viral sequences used in the structure prediction for the Phold database @bourasProteinStructureinformedBacteriophage2026.

### Foldseek
Foldseek [@vankempenFastAccurateProtein2024] converts predicted protein structures into 3Di sequences and subsequently performs fast structure similarity searches using these 3Di sequences. Here, we used `foldseek easy-search` (Foldeek v10.941cd33) with default settings to search the predicted structures of our test set against the BFVD.

### Reseek
Reseek [@edgarProteinStructureAlignment2024] aligns structures in a similar way to Foldseek, but uses a different structural alphabet consisting of ~10^11^ states, promising better sensitivity for remote homologs. 

We used an adapted version of Reseek v2.7 in sensitive mode that switches the loading of the queries and database to get more accurate E-value calculations (see [this discussion](https://github.com/rcedgar/reseek/issues/18)). 

## Functional category assignment
To functionally annotate our test proteins for each method, weights in form of the negative log E-value (or a weight of 1000 in case of an E-value of 0) were given to each hit. These weights were summed per functional category (assigned to each entry in the BFVD; see [above](#functional-classifier)) and the category with the largest weight was chosen as the functional category for each test protein. Informative categories were prioritized over uninformative categories (ie. "Hypothetical Protein", "unknown", "Other function") when the weight of the informative category was at least 50% the weight of the uninformative categories.

## Computational benchmarking
All steps in our analysis were benchmarked for both time and memory usage using hyperfine v1.20.0 [@Peter_hyperfine_2023]. GPU memory consumption was measured by sampling usage every 200 ms with nvidia-smi v580.82.07, recording the peak value observed after each step.

All CPU analysis steps were performed on an HPC system using one IceLake node with 72 cores (2 Intel Xeon Platinum 8360Y CPUs@2.4 GHz; 256 GiB RAM), except for the MSA generation step which was performed on an IceLake big memory node with 72 cores (2 Intel Xeon Platinum 8360Y CPUs@2.4 GHz; 2048 GiB RAM). The GPU analyses were performed on one NVIDIA H100 SXM5 GPU (80 GiB HBM3), mounted on a node equipped with 2 AMD EPYC 9334 CPUs (768 GiB RAM).

## Code availibility
Code to reproduce this analysis is available on [GitHub](https://github.com/LanderDC/annotation_benchmark/).

## Data availibility
[upload to zenodo]{.mark}

# Acknowledgements
LDC acknowledges support from the Research Foundation - Flanders (grants 11L1325N and V451025N). The computational resources and services used in this work were provided by the VSC (Flemish Supercomputer Center), funded by the Research Foundation - Flanders (FWO) and the Flemish Government.

# Declaration of LLM usage
During the preparation of this manuscript, the authors made use of large language model (LLM)-based artificial intelligence tools to assist with literature searches, restructuring and improving sentences, and generating code for data analysis. All AI-generated content and code was critically reviewed, verified, and edited by the authors, and full responsibility for the accuracy and integrity of the work rests with the authors alone.

# References
::: {#refs}
:::

{{< pagebreak >}}

# Supplementary Figures
![The distribution of the number of genomes from the different viral Kingdoms in the test set.](../figures/data_info/kingdom_distribution.pdf){#suppfig-kingdom fig-pos='H'}

![The number of test proteins with at least 1 hit for each method across different E-value thresholds.](../figures/search_results/unique_queries_vs_evalue.pdf){#suppfig-query-vs-evalue fig-pos='H'}

![Pairwise comparisons of shared hits between search methods for (A) the best 25 hits, (B) the best 25 hits after filtering on E-value (E < 1e-03), and (C) the top hits for each query.](../figures/search_results/hit_overlap_methods.pdf){#suppfig-pairwise-method fig-pos='H'}

![Workflow runtime increase compared to the fastest workflow (MMseqs2).](../figures/computational_benchmark/time_fold_increase.pdf){#suppfig-time-increase fig-pos='H'}

![The number of informative proteins in BFVD by their main function, colored by eukaryotic or bacteriophage protein.](../figures/data_info/BFVD_informative_classification.pdf){#suppfig-bfvd-classification fig-pos='H'}

::: {#suppfig-llm-diagram}
```{mermaid}
%%| fig-width: 2.5
%%| fig-height: 9
---
config:
  theme: neutral
  themeVariables:
    fontSize: 12
---
flowchart TD
    A([🚀 Start]) --> B[Read command-line arguments</br>proteins, categories, output, model, etc.]
    B --> C[Load proteins JSON file]
    C --> D[Load categories JSON file</br>and normalize into a clean list]
    D --> E[Load AI language model</br>from HuggingFace]

    E --> F{Optional:</br>Quantize model?}
    F -- 4-bit or 8-bit --> G[Apply memory compression</br>to reduce GPU usage]
    F -- No --> H[Load model at full precision]
    G --> I[Model ready]
    H --> I

    I --> J{Does an output file</br>already exist?}
    J -- Yes --> K[Load existing results</br>Skip already-processed proteins]
    J -- No --> L[Start fresh]
    K --> M
    L --> M

    M([🔁 Loop over each protein]) --> N{Already</br>processed?}
    N -- Yes --> M
    N -- No --> O[Extract protein names</br>and virus taxonomy]

    O --> P[Build prompt for AI:</br>protein names + taxonomy +</br>predefined categories]
    P --> Q[Send prompt to</br>AI language model]
    Q --> R[Receive AI response</br>in JSON format]

    R --> S{Response</br>parseable?}
    S -- Yes --> T[Validate assigned categories</br>against predefined list]
    S -- No --> U[⚠️ Log warning</br>Store empty categories]

    T --> V{Any categories</br>need repair?}
    V -- Yes --> W[Auto-fix minor name issues</br>e.g. missing parentheses]
    V -- No --> X
    W --> X[Remove any invalid categories]

    X --> Y[Save result for this protein]
    U --> Y
    Y --> Z{Checkpoint:</br>Every N proteins}
    Z -- Yes --> AA[💾 Save progress to output file]
    Z -- No --> M
    AA --> M

    M -- All proteins done --> AB[💾 Final save to output JSON]
    AB --> AC{Were any original</br>labels provided?}
    AC -- Yes --> AD[Compute accuracy metrics</br>Precision / Recall / F1]
    AC -- No --> AE
    AD --> AE([✅ Done!</br>Results saved to output file])

    style A fill:#4CAF50,color:#fff
    style AE fill:#4CAF50,color:#fff
    style M fill:#2196F3,color:#fff
    style Q fill:#9C27B0,color:#fff
    style AA fill:#FF9800,color:#fff
    style AB fill:#FF9800,color:#fff
    style U fill:#f44336,color:#fff
```

Flowchart of the LLM classification pipeline for eukaryotic viruses based on the protein names in databases like GenBank, UniProt, etc.
:::

```{=latex}
\clearpage
\newgeometry{left=1.3cm,right=1.3cm,top=0cm,bottom=0cm}
```

::: {#suppfig-method-diagram}

```{mermaid}
%%| fig-width: 7
%%| fig-height: 9
%%| out-width: 100%
---
config:
  theme: neutral
  themeVariables:
    fontSize: 12
---
flowchart TD
    A[Goal<br/>Annotate proteins from eukaryotic viruses] --> B[Test set selection from ICTV VMR v40.2]
    B --> B1[Keep entries with no bacteria/archaea as host<br/>1,054 entries; 1,000 species]
    B1 --> B2[Download associated proteins from GenBank<br/>11,360 proteins]
    B2 --> B3[Split proteins longer than 2,500 aa<br/>for structure-prediction memory limits]
    B3 --> B4[Final test set: 11,467 protein sequences]

    A --> C[Reference database: BFVD]
    C --> C1[BFVD contains viral proteins + predicted structures]
    C1 --> C2[Map UniProt accessions to UniParc when needed]
    C2 --> C3[Assign functional categories to BFVD entries]
    C3 --> C4[Eukaryotic-virus proteins: custom name-based classifier]
    C3 --> C5[Bacteriophage proteins: Empathi PHROG-like classes]

    B4 --> D[Run homology searches against BFVD]
    C4 --> D
    C5 --> D

    D --> E{Method family}

    E --> E1[Sequence methods]
    E1 --> E11[BLASTp]
    E1 --> E12[DIAMOND blastp sensitive]
    E1 --> E13[MMseqs2 easy-search]

    E --> F1[Embedding methods]
    F1 --> F11[ProstT5: protein sequence -> 3Di sequence]
    F11 --> F12[Search with Foldseek easy-search]
    F1 --> F21[TEA: protein sequence -> TEA sequence using ESM2]
    F21 --> F22[Search with MMseqs2 + TEA scoring settings]

    E --> G1[Structure methods]
    G1 --> G11[Predict structures with Boltz-2]
    G11 --> G12[MSA from adapted ColabFold + viral additions]
    G12 --> G13[Search predicted structures with Foldseek]
    G12 --> G14[Search predicted structures with adapted Reseek]

    E11 --> H[Per-method hit list for each test protein]
    E12 --> H
    E13 --> H
    F12 --> H
    F22 --> H
    G13 --> H
    G14 --> H

    H --> I[Filter hits by E-value < 1e-03 for category assignment]
    I --> J["Convert each hit to a category score<br/>weight = -log(E-value), or 1000 when E-value = 0"]
    J --> K[Sum scores per category for each test protein]
    K --> L[Choose category with highest total score]
    L --> M[Priority rule: if best informative category >= 50% of best uninformative score,<br/>keep informative category]
    M --> N[Final functional assignment per test protein, per method]

    subgraph U[Uninformative categories]
      U1[Hypothetical Protein]
      U2[unknown]
      U3[Other function]
    end

    M -. uses .-> U
```

Flowchart of the annotation pipeline used in this study. Eukaryotic virus proteins were selected from ICTV VMR v40.2, filtered to exclude bacteria- and archaea-hosted entries, and expanded into a final test set after retrieving all associated protein sequences and splitting very long proteins for structure prediction. The BFVD reference database was prepared separately by assigning functional categories to its entries using a custom name-based classifier for eukaryotic virus proteins and PHROG-like categories for bacteriophage proteins by Empathi. The test proteins were then compared against BFVD using three sequence methods (BLASTp, DIAMOND, MMseqs2), two embedding-based methods (ProstT5-3Di and TEA), and two structure-based methods (Foldseek and Reseek after Boltz-2 structure prediction). Hits were filtered by E-value, converted into weighted category scores, summed per functional category, and the highest-scoring category was chosen for each test protein, with a rule that preserves informative categories when they are sufficiently supported over uninformative ones.
:::

```{=latex}
\restoregeometry
\clearpage
```

```{=latex}
\FloatBarrier
```

{{< pagebreak >}}

```{=latex}
\clearpage
\newgeometry{left=1cm,right=1cm,top=2cm,bottom=2cm}
```
# Supplementary Tables
```{r}
#| label: tbl-supp-functional-categories
#| tbl-cap: "Final eukaryotic virus functional categories used for annotation, including their activity and process they are involved in."
#| warning: false
#| message: false
functional_categories_path <- "../data/viralzone_protein_categories.json"
functional_categories <- jsonlite::fromJSON(functional_categories_path)

functional_categories_tbl <- functional_categories |>
  dplyr::select(
    process,
    Activity,
    `Enzyme/Protein`
  ) |>
  dplyr::rename(Process = process) |>
  dplyr::arrange(Process)

knitr::kable(
  functional_categories_tbl,
  format = if (knitr::is_latex_output()) "latex" else "pipe",
  booktabs = TRUE,
  longtable = knitr::is_latex_output(),
  linesep = ""
)
```

```{=latex}
\restoregeometry
\clearpage
```